Flights Data for November and December 2013
========================================================
Team: Sam Helmich, Sen Zhou, Jennifer Chang

Date: 20 February 2014

Getting Ready
-----------------
Load required libraries
```{r, message=FALSE}
library(RSQLite)
library(RSQLite.extfuns)
library(dplyr)
library(plyr)
```

Create a SQLite database flights in your working directory
```{r}
my_db <- src_sqlite("flights", create = T)
```

Get the Data
------------------
Download dataset for November and December 2014 from the following website.  
http://www.transtats.bts.gov/DL_SelectFields.asp?Table_ID=236&DB_Short_Name=On-Time

Select the following variables if they are not already selected: DAY_OF_WEEK,ORIGIN_AIRPORT_ID, DEST_AIRPORT_ID, DEP_TIME, DEP_DELAY, ARR_TIME, ARR_DELAY, CANCELLED, CANCELLATION_CODE

Load the data into the SQL database.
```{r}
nov<-read.csv("laxnov.csv")
dec<-read.csv("laxdec.csv")
flights <-rbind(nov,dec)
hflights_sqlite <- copy_to(my_db, flights)
dim(hflights_sqlite)
rm(flights,nov,dec)
```

LAX Airport Delays
-------------------
We decided to focus on LAX (Los Angeles International) airport.  
LAX ID = 12892

We looked at departure and arrival delays and how it was related to days of the week and time.
```{r fig.width=12, fig.height=5}
lax <- hflights_sqlite %.%
          filter(ORIGIN_AIRPORT_ID == 12892 | DEST_AIRPORT_ID == 12892) %.%
          group_by(DAY_OF_WEEK) %.%
          select(DAY_OF_WEEK, DEP_DELAY, DEP_TIME, ARR_DELAY, ARR_TIME)

lax.df <- as.data.frame(lax)

lax.df$DAY_OF_WEEK <- factor(lax.df$DAY_OF_WEEK)
lax.df$DAY_OF_WEEK <- mapvalues(lax.df$DAY_OF_WEEK, from = c("1", "2", "3", "4", "5", "6", "7"), to = c("Sun", "Mon", "Tues", "Wed", "Thurs", "Fri", "Sat"))

library(ggplot2)
qplot(DEP_TIME, DEP_DELAY, data = lax.df, facets = .~DAY_OF_WEEK, main = "Departures from LAX", ylab="Delay in Minutes", xlab="Time")+theme(axis.text.x = element_text(angle = 45, hjust = 1))
qplot(ARR_TIME, ARR_DELAY, data = lax.df, facets = .~DAY_OF_WEEK, main = "Arrivals to LAX", ylab="Delay in Minutes", xlab="Time")+theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
There were more delays in the evenings to nights.  The lack of delays in the early mornings were probably due the airport shutting down at night.

LAX Airport Cancellations
-------------------
We plotted cancellations based on distance for LAX arrivals, departures, and both. 
```{r}
lax.all <- filter(hflights_sqlite, DEST_AIRPORT_ID == 12892 | ORIGIN_AIRPORT_ID == 12892)
lax.canc<- filter(lax.all, CANCELLED==1)
lax.canc.df <-collect(lax.canc)
qplot(DISTANCE, data=lax.canc.df, main="Cancelled Flights from or to LAX")

# arrivals
lax.canc.arr<-filter(lax.canc.df,DEST_AIRPORT_ID == 12892)
qplot(DISTANCE, data=lax.canc.arr,main = "Cancelled Flights to LAX")
lax.canc.dep<-filter(lax.canc.df,ORIGIN_AIRPORT_ID == 12892)
qplot(DISTANCE, data=lax.canc.dep,main = "Cancelled Flights from LAX")
```
For the most part, there were more short distance flight cancellations than long distance flight cancelations.
We looked at the CANCELLATION_CODE in order to determine the reason for cancelation however most of the values were missing.
```{r}
lax.canc.df$CANCELLATION_CODE <- mapvalues(lax.canc.df$CANCELLATION_CODE, from = c("A", "B", "C", "D"), to = c("Carrier", "Weather", "National Air System", "Security"))
qplot(CANCELLATION_CODE, data=lax.canc.df)+theme(axis.text.x = element_text(angle = 30, hjust = 1,size=14))
sum(lax.canc.df$CANCELLATION_CODE=="Carrier")
sum(lax.canc.df$CANCELLATION_CODE=="Weather")
sum(lax.canc.df$CANCELLATION_CODE=="National Air System") 
sum(lax.canc.df$CANCELLATION_CODE=="Security")
```

<!-- Comment Chunk
Code for Sam Helmich
//```{r}
library(RSQLite)
library(dplyr)

## Variables in dataset:
# [1] "DAY_OF_WEEK"           "ORIGIN_AIRPORT_ID"     "ORIGIN_AIRPORT_SEQ_ID" "ORIGIN_CITY_MARKET_ID"
#  [5] "DEST_AIRPORT_ID"       "DEST_AIRPORT_SEQ_ID"   "DEST_CITY_MARKET_ID"   "DEP_TIME"             
#  [9] "DEP_DELAY"             "ARR_TIME"              "ARR_DELAY"             "X"  

# This part runs
my_db <- src_sqlite("my_db.sqlite3", create = T)
nov <- read.csv("november.csv")
dec <- read.csv("december.csv")
data <- rbind(nov, dec)
hflights_sqlite <- copy_to(my_db, data, temporary = FALSE)
rm(nov, dec, data)

tbl_vars(hflights_sqlite)

# We'll go with LAX
# LAX ID = 12892
lax <- hflights_sqlite %.%
          filter(ORIGIN_AIRPORT_ID == 12892 | DEST_AIRPORT_ID == 12892) %.%
          group_by(DAY_OF_WEEK) %.%
          select(DAY_OF_WEEK, DEP_DELAY, DEP_TIME, ARR_DELAY, ARR_TIME)

lax.df <- as.data.frame(lax)
          
lax.df$sch.depart <- with(lax, DEP_TIME - DEP_DELAY)
lax.df$sch.arr <- with(lax, ARR_TIME - ARR_DELAY)
library(ggplot2)

qplot(DEP_TIME,DEP_DELAY, data = lax.df, facets = .~DAY_OF_WEEK, main = "Departures")
qplot(ARR_TIME, ARR_DELAY, data = lax.df, facets = .~DAY_OF_WEEK, main = "Arrivals")

```

Code for Jennifer Chang
//```{r fig.width=7, fig.height=6}
#data_loc<-"/Users/jenchang/Desktop/STAT585/Projects/Proj6/"

# install & load required packages
install.packages('dplyr')
install.packages('RSQLite')
install.packages("RSQLite.extfuns")
library(dplyr)
library(RSQLite)
library(RSQLite.extfuns)


# Load Data
nov <- read.csv("/Users/jenchang/Desktop/STAT585/Projects/Proj6/data/november2013.csv")
dec <- read.csv("/Users/jenchang/Desktop/STAT585/Projects/Proj6/december2013.csv")

# connect to database and move data over
my_db <- src_sqlite("my_db.sqlite3", create = T)
hflights_sqlite <- copy_to(my_db, nov, temporary = FALSE)

# focus on LAX 12892
LAX_flights<-filter(hflights_sqlite,ORIGIN_AIRPORT_ID == 12892 | DEST_AIRPORT_ID == 12892)
out_df <-collect(LAX_flights)
dep<-filter(hflights_sqlite,ORIGIN_AIRPORT_ID==12892)
dep_df<-collect(dep)
arr<-filter(hflights_sqlite,DEST_AIRPORT_ID==12892)
arr_df<-collect(arr)

canceled<-filter(hflights_sqlite, ORIGIN_AIRPORT_ID==12892, CANCELLED==1)
can_df<-collect(canceled)
head(CANCELLATION)
# Plot
library(ggplot2)
qplot(DEP_TIME, DEP_DELAY, data=dep_df, colour=factor(DAY_OF_WEEK))
qplot(DEP_TIME, DAY_OF_WEEK, data=dep_df)

qplot(CANCELLED, DISTANCE, data=dep_df)
qplot(CANCELLED, DISTANCE, data=arr_df)
qplot(DISTANCE, CANCELLED, data=out_df)
qplot(DISTANCE, data=out_df, fill=factor(CANCELLED))

dim(can_df$CANCELLATION_CODE)

# create table
dbGetQuery(co, "CREATE TABLE IF NOT EXISTS nd_flights (
  obs INT PRIMARY KEY NOT NULL,
  totbill DOUBLE,
  tip DOUBLE,
  sex CHAR(1),
  smoker CHAR(3),
  day CHAR(3),
  time CHAR(5),
  size INT
  )")

# open sqlite connection
my_db <- src_sqlite(“mydb.sqlite3”, create=T)

# load your data into database & setting indexes
yourdata_sqlite <- copy_to(my_db, yourdata, temporary=FALSE, indexes = list( c(“col1”, “col2”), col3))
yourdata_sqlite_df <- tbl(my_db, sql(“SELECT * from yourdata”))  # do we need this?

# ===  Basic Actions (VERBS) === looks the same as above!!!! 
filter( yourdata_sqlite_df, col1==”value1”, col2==”value2”) 
arrange( yourdata_sqlite_df, col2, col4)
select( yourdata_sqlite_df, col2,col4) 
mutate( yourdata_sqlite_df, diff = col2 - col4, percentage = col2 / col4)
summarize( yourdata_sqlite_df, avg = mean(col2, na.rm=TRUE) 

# get tbl_df of results using collect (sql statements not executed on database until R needs to print it out)
out <- select( yourdata_sqlite_df, col2,col4)
your_final_data_df <-collect(out)

# close sqlite connection
dbDisconnect(my_db)

```
-->

